Score-based tests of measurement invariance: Use in practice

### 2013-08-23 frontiers/frontiers.Rnw
Frontiers ms, additions to tutorial section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@241 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

c(dm$p.value, cvm$p.value, maxlm$p.value)
@
In the code above, the \verb+parm=1:4+ argument contains parameter numbers reflecting the factor loadings that we wish to test.  The ordering of parameters can be obtained via the command \verb+coef(reduced)+.  Two of the three $p$-values output at the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).

%% Aug 21: Ed stopped here.  TODO: graphs

\begin{figure}
\caption{Get Continuous Estimate DM Alternative Way}
\label{fig:conti1res}
\setkeys{Gin}{width=\textwidth}
<<conti1res, fig=TRUE, height=7, width=7>>=
DM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
DM
@
\end{figure}

\begin{figure}
\caption{Get Continuous Estimate CvM Alternative Way}
\label{fig:conti2res}
\setkeys{Gin}{width=\textwidth}
<<conti2res, fig=TRUE, height=7, width=7>>=
CvM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
CvM
@
\end{figure}

<<conti3resmaxLM>>=
maxLM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
maxLM
@
Since there are more than one observations located in each age level. 
So the observation order within each age level is not unique, thus, 
maxLM can't produce the plot. Actually, all three continuous statistics 
were calculated on the arbitrary order that particular dataset entered 
in each age level. In order to accurately account for the multiple observations 
at the same age level, it is more appropriate to use the proposed ordinal test.

\subsection{Ordinal Test}
As a matter of fact, there is only a partial ordering of individuals 
with respect to $V$,i.e. observations with the same level 
of $V$ have no unique ordering. The ordinal statistics 
proposed here are similar to those described in Equation 
and above, except that we focus on "bins" of individuals 
at each level of the ordinal variable.

The difference between the continuous test and ordinal 
test is that we don't force arbitrary order on the observations 
within the same age groups. Therefore, this ordinal test should 
reflect more real status of the data. In terms of code, we still need the gefp 
function to get the B estimate. After that, we can use different 
functions to aggregate and scale B estimation. 
The function are called ordwmax and ol2bb1.fun, respectively. 
<<Ordinal Test>>=
gefp1<-gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval.ord=rep(NA,2)
pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value

if(file.exists("opval1.rda")){
  load("opval1.rda")
} else {
  opval1 <- ordL2BB(gefp1)
  save(opval1, file="opval1.rda")
}
pval.ord[2]<-sctest(gefp1, functional=opval1)$p.value
pval.ord
@

Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
\begin{figure}
\caption{Get Ordinal Estimate WDMo Alternative Way}
\label{fig:ord1res}
\setkeys{Gin}{width=\textwidth}
<<ord1res, fig=TRUE, height=7, width=7>>=
WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
WDMo 
@
\end{figure}

\begin{figure}
\caption{Get Ordinal Estimate maxLMo Alternative Way}
\label{fig:ord2res}
\setkeys{Gin}{width=\textwidth}
<<ord2res, fig=TRUE, height=7, width=7>>=
maxLMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
               fnl=opval1, fplot=TRUE)
maxLMo 
@
\end{figure}

In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. 
Both $p-$values are clearly larger than that of the 
likelihood ratio test and neither is significant at $\alpha=0.05$. 
This result is consistent with alternative fit indices used in Froh et al paper\citeA{FroFan11}.

>>>>

round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
@
In the code above, the \verb+parm=1:4+ argument means that we wish to test model parameter numbers 1 through 4 (which are the factor loadings), which are supplied by lavaan.  The ordering of parameters can be obtained via the command \verb+coef(reduced)+.  We see that two of the three $p$-values output at the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).

%% TODO Illustrate with graphs?

%% THE FOLLOWING HAS BEEN MOVED
%% \begin{figure}
%% \caption{Get Continuous Estimate DM Alternative Way}
%% \label{fig:conti1res}
%% \setkeys{Gin}{width=\textwidth}
%% <<conti1res, fig=TRUE, height=7, width=7>>=
%% DM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
%% DM
%% @
%% \end{figure}

%% \begin{figure}
%% \caption{Get Continuous Estimate CvM Alternative Way}
%% \label{fig:conti2res}
%% \setkeys{Gin}{width=\textwidth}
%% <<conti2res, fig=TRUE, height=7, width=7>>=
%% CvM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
%% CvM
%% @
%% \end{figure}

%% <<conti3resmaxLM>>=
%% maxLM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
%% maxLM
%% @
The tests carried out here assume a unique ordering of individuals by age, but this is obviously not the case.  To compute the statistics and $p$-values, the strucchange package implicitly assumed an arbitrary ordering of individuals who are tied on age.  If we were to change this ordering, the resulting statistics and $p$-values would also change, which is obviously problematic.  To accurately account for the multiple observations  
at the same age level, we must used the ordinal tests from Equations~\eqref{eq:wdm} and~\eqref{eq:maxlmo}.  These are described next.

\subsection{Ordinal Treatment}

%% As a matter of fact, there is only a partial ordering of individuals 
%% with respect to $V$,i.e. observations with the same level 
%% of $V$ have no unique ordering. The ordinal statistics 
%% proposed here are similar to those described in Equation 
%% and above, except that we focus on "bins" of individuals 
%% at each level of the ordinal variable.

The difference between the ordinal test statistics and their
continuous counterparts is that the ordinal statistics do not require
an ordering of individuals  
within the same age group.  To compute the test statistics, we allow
the scores of all tied individuals to enter the cumulative sum
(Equation~\eqref{eq:cumscore}) simultaneously.  This results in
modified critical values and test statistics that are sensitive to
measurement invariance violations that are monotonic with age group.
%% Therefore, this ordinal test should 
%% reflect more real status of the data. In terms of code, we still need the gefp 
%% function to get the B estimate. After that, we can use different 
%% functions to aggregate and scale B estimation. 
%% The function are called ordwmax and ol2bb1.fun, respectively.

To carry out the tests, we can rely on the same function that we used
for the continuous test statistics.  As mentioned previously,
calculation of the ``maxLMo'' statistic (Equation~\eqref{eq:maxlmo})
can be lengthy from the need to simulate critical values.
<<critvals,echo=FALSE>>=
if(file.exists("opval1.rda")){
  load("opval1.rda")
} else {
  opval1 <- ordL2BB(gefp1)
  save(opval1, file="opval1.rda")
}
## Get test statistics using fnl=opval1 without showing it in the paper
wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo",
                 fnl=opval1)
@

<<ordtest,eval=FALSE>>=
wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo")
@ 
<<pvals>>=
round(c(wdmo$p.value, maxlmo$p.value), 2)
@

%% TODO Describe graphs

%% gefp1<-gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
%%        vcov=info_full,sandwich=FALSE,parm=1:4)
%% pval.ord=rep(NA,2)
%% pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value

%% pval.ord[2]<-sctest(gefp1, functional=opval1)$p.value
%% pval.ord
%% @

%Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
\begin{figure}
\caption{Get Ordinal Estimate WDMo Alternative Way}
\label{fig:ord1res}
\setkeys{Gin}{width=\textwidth}
<<ord1res, fig=TRUE, height=7, width=7, echo=FALSE>>=
WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
#WDMo 
@
\end{figure}

\begin{figure}
\caption{Get Ordinal Estimate maxLMo Alternative Way}
\label{fig:ord2res}
\setkeys{Gin}{width=\textwidth}
<<ord2res, fig=TRUE, height=7, width=7, echo=FALSE>>=
maxLMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
               fnl=opval1, fplot=TRUE)
#maxLMo 
@
\end{figure}

In computing the ordinal test statistics, we
obtain~$p=\Sexpr{round(wdmo$p.value,2)}$
and~$p=\Sexpr{round(maxlmo$p.value,2)}$, respectively.
Both $p-$values are clearly larger than that of the 
likelihood ratio test and neither is significant at $\alpha=0.05$. 
This provides evidence that there is no measurement
invariance violation that is monotonic with age group.  Instead, given
the large sample size, the likelihood ratio test may be overly
sensitive to anomalous, non-monotonic violations at one (or a few) age groups.

In the above sections, we have illustrated the score-based tests'
computation in R.  We suspect that the ordinal tests will be most
popular with users, because measurement invariance tests are typically
carried out across categories (ordered or not), as opposed to
continuous variables.  Thus, in the sections below, we report
novel simulations to study the ordinal statistics' expected behavior
in practice.  In particular, we wish to study (i) the extent to which the
ordinal statistics attribute measurement invariance violations to the
correct parameter(s), and (ii) the extent to which the tests are
robust to model misspecification.
